---
title: "Reproducible Research: Peer Assessment 1"
author: "José Alberto Hernández"
resource_files:
 - activity.csv
output: 
  html_document:
    keep_md: yes
    theme: flatly
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "figure/"
)
```

## Loading and preprocessing the data

Input data variables
```{r}
url_data <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zip_data <- "activity.zip"
csv_data <- "activity.csv"
```

[Download the zip]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip") file containing activity data, save it to the working directory
```{r}
download.file(url_data, destfile = zip_data)
```

Unzip the zip file to get the csv file
```{r}
unzip(zip_data)
```

Read the activity data from the csv file considering:

- (,) Comma as delimiter
- (") Quote as quote indicator
- NA string as missing values

```{r}
data <- read.csv(csv_data, sep = ",", quote = "\"", na.strings = "NA")
```

Exploring the raw data
```{r}
str(data)
```

We can see

- `r format(nrow(data), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` observations
- `r ncol(data)` variables `r names(data)`
- The date variable is `r class(data$date)` class
- There are missing values (NA) in the steps variable

```{r}
summary(data)
```
There are `r format(sum(is.na(data$steps)), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` NA's

```{r}
head(data)
```

Cleaning the raw data, setting the date as Date with the format YYYY-MM-DD
```{r}
# Convert the date variable to Date type 
data$date <- as.Date(as.character(data$date), "%Y-%m-%d")
str(data)
```
Now, the date variable is `r class(data$date)` class

## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day without missing values
```{r}
# Calculate the total number of steps by day without missing values
total.steps <- tapply(data$steps, data$date, FUN = sum, na.rm = TRUE)
```

Plot (Histogram) of the total number of steps walked each day
```{r histogram-no-missing-values}
# Create the Histogram to view the steps by day with filled data
hist(total.steps, xlab = "Steps", main = "Histogram steps by day", 
     breaks = 25, col = 'grey', cex.main = 0.9)

# mean and median values round with 2 decimals
total.steps.mean = round(mean(total.steps, na.rm = TRUE), 2)
total.steps.median = round(median(total.steps, na.rm = TRUE), 2)

# Print the mean and median lines
abline(v = total.steps.mean, lwd = 3, col = 'red')
abline(v = total.steps.median, lwd = 3, col = 'green')

# Print the legend with the mean and median values
legend('topright', lty = 1, lwd = 3, col = c("red", "green"), cex = .8, 
               legend = c(paste('Mean: ', total.steps.mean),
               paste('Median: ', total.steps.median)))
```

*mean* and *median* functions without missing values
```{r}
mean(total.steps, na.rm = TRUE)
median(total.steps, na.rm = TRUE)
```
The *mean* is `r format(round(total.steps.mean,2), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` and the *median* is `r format(total.steps.median, digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")`

## What is the average daily activity pattern?
Agregate the data by date without the missing values
```{r}
# Agregate the dataset to get the mean by interval withou the missing values
average.by.day <- aggregate(x = list(steps = data$steps), 
                      by = list(interval = data$interval), FUN = mean, na.rm = TRUE)
```

Time series plot with

- (x-axis) as the 5-minute interval, and 
- (y-axis) as the average number of steps taken, averaged across all days

```{r timeseries-average-steps-interval}
# Plot the time series
with(average.by.day, {
        plot(interval, steps, type = 'l',
             main = 'Average steps by interval',
             xlab = '5-minute interval', ylab = 'Average steps')
        })

# Find the maximum of step and interval
maximum = average.by.day[which.max(average.by.day$steps), ]

# The string label for the maximum of step and interval
max_lab = paste('Maximum of ', round(maximum$steps, 2), ' steps in the ', maximum$interval, 'th interval', sep = '')

# Print the point with the maximum of steps and interval
points(maximum$interval, maximum$steps, col = 'blue', lwd = 3, pch = 19)

# Print the label with the maximum number of steps and interval
legend("topright", legend = max_lab, bty = 'n')
```

#### Which is tne maximum number of steps in the 5-minute interval?
```{r}
average.by.day[which.max(average.by.day$steps), ]
```
> `r max_lab` 

## Imputing missing values
As we saw with the *summary* function, we have `r format(sum(is.na(data$steps)), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` observations with missing values (coded as NA). The presence of missing values may introduce bias into some calculations or summaries of the data.
```{r}
sum(is.na(data$steps))
```

Function for set the *mean* value of 5 minute interval where there are missing values
```{r}
fill.value <- function(steps, interval) {
  filled <- NA
  if (!is.na(steps)) 
    filled <- c(steps) 
  else 
    filled <- (average.by.day[average.by.day$interval == interval, "steps"])
  return(filled)
}
```

Replacing each of these `r format(sum(is.na(data$steps)), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` missing values, with the *mean* value of its 5 minute interval
```{r}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```

Validate there is not more missing values in the *filled.data* dataset
```{r}
sum(is.na(filled.data$steps))
```
Now, there are `r format(sum(is.na(filled.data$steps)), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` NA's in the *filled.data* dataset

Calculate the total number of steps taken per day with all the observations
```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN = sum)
```

Plot (Histogram) of the total number of steps taken each day, with all the observations
```{r histogram-filled-data}
# Create the Histogram to view the steps by day with filled data 
hist(total.steps, xlab = "Steps", main = "Histogram steps by day with filled data", 
     breaks = 25, col = 'grey', cex.main = 0.9)

# mean and median values round with 2 decimals
total.steps.mean = round(mean(total.steps), 2)
total.steps.median = round(median(total.steps), 2)

# Print the mean and median lines
abline(v = total.steps.mean, lwd = 3, col = 'red')
abline(v = total.steps.median, lwd = 3, col = 'green')

# Print the legend with the mean and median values
legend('topright', lty = 1, lwd = 3, col = c("red", "green"), cex = .8, 
               legend = c(paste('Mean: ', total.steps.mean),
               paste('Median: ', total.steps.median)))
```

*mean* and *median* functions with all the observations
```{r}
mean(total.steps)
median(total.steps)
```
The *mean* and *median* are equal to `r format(round(mean(total.steps),2), digits=9, decimal.mark=".",small.mark=",", small.interval=3, big.mark=",")` 

## Are there differences in activity patterns between weekdays and weekends?

Function for getting the day type (weekday / weekend) based in the date value passed as parameter
```{r}
weekday.or.weekend <- function(dayname) {
  if (dayname %in% c("Saturday", "Sunday")) {
    return("Weekend") 
  }
  return("Weekday") 
}
```

Set the day variable with the *weekday.or.weekend* function
```{r}
# Set the dayname variable to set: day of week
filled.data$dayname = weekdays(filled.data$date)

# Add daytype variable to set: Weekday or Weekend
filled.data$daytype = as.factor(apply(as.matrix(filled.data$dayname), 1, weekday.or.weekend))
```
Now we have `r ncol(filled.data)` variables `r names(filled.data)` in the *filled.data* dataset

Agregate the *filled.data* dataset in the steps variable agains the interval and day variables
```{r}
# Summarize the dataset with the mean grouped by interval and daytype
library(data.table)
summary.by.daytype = data.table(filled.data)[, list(avg_steps = mean(steps)), 
                          by = list(interval, daytype)]

```

Panel plot for comparing the difference in activity patterns between weekday and weekends
```{r weekday-weekend-patterns}
library(lattice)
# XY Plot for the relation with the avg_steps and interval variables, splited by daytype variable
xyplot(avg_steps ~ interval | daytype, data = summary.by.daytype,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Steps',
      layout = c(1,2))
```

> **Yes!**, there are differences, the group of enthusiasts walks more in the Weekends, but the group of enthusiasts walks more, early in the morning around 8.00 hrs, in the Weekdays